Model-based Clustering for Neural Populations

### 2021-10-03 documents/cluster_neural_population.tex
cluster_initial_done

%\section{Model Check [Problems for \(\mathbf{b}\) and \(\mathbf{A}\) partially solved]} \label{problems}
%
%\subsection{Estimation of \(\mathbf{d}\) and \(\mathbf{C}\) alone}
%First, turn off all others except for loading (related) parameters and set them to be true values. There are 2 versions: (1) update priors as in subsection \ref{loading prior} and (2) no update of priors, and just set the prior for each as the standard normal.
%For the estimation with prior update, the results are mean from iteration 50 to 100 (Figure \ref{loading alone}).
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image016.png}}%
%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image017.png}}%
%	}
%	\caption{estimation of loading}
%	\label{loading alone}
%\end{figure}
%
%The mixing of convergence for loading priors (in subsection \ref{loading prior}) are fine. The update of priors don't influence results a lot, but this will help clustering a lot. It seems estimation of loading is fine. Let's see what happens when estimation of latents \(\mathbf{x}_t\) is on at the same time.
%
%\subsection{Estimation of \(\mathbf{d}\), \(\mathbf{C}\) and \(\mathbf{x}_{t}\)}
%When the estimation of \(\mathbf{x}_t\) is on, we need more iterations. The results are mean from iteration 500 to 1000 (Figure \ref{loading and latent}).
%
%The estimation of loading:
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image018.png}}%
%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image020.png}}%
%	}
%	\caption{estimation of loading, with latents on}
%	\label{loading and latent}
%\end{figure}
%
%And the estimation of corresponding latents (Figure \ref{loading and latents}):
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[no update of prior]{\includegraphics[width=0.6\textwidth]{image019.png}}%
%		\subfigure[with update of prior]{\includegraphics[width=0.6\textwidth]{image021.png}}%
%	}
%	\caption{estimation of loading}
%	\label{loading and latents}
%\end{figure}
%
%Things are still fine. The remaining parts may influence latent estimations are dynamics (\(\mathbf{b}\), \(\mathbf{A}\)) and prior process noise \(\mathbf{Q}\).
%
%\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\), \(\mathbf{Q}\) and \(\mathbf{x}_t\)}
%I show the results of block-diagonal (Figure \ref{bAQX_blkDiag}) and full (Figure \ref{bAQX_full}) \(\mathbf{Q}\). These results are from 2000 MCMC samples. The latent and dynamics are from average from 500 to 2000 iterations.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_trace.png}}%
%		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_latent.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_A.png}}%
%	}
%	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, block-diagonal process noise}
%	\label{bAQX_blkDiag}
%\end{figure}
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_full_trace.png}}%
%		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_full_latent.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_full_A.png}}%
%	}
%	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, full process noise}
%	\label{bAQX_full}
%\end{figure}
%
%Both block-diagonal and full looks fine, but the mixing of  full process noise version looks better.

@@

%There are two set of simulation examples. The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics. In all following results, I fit things with both \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{block-diagonal} and \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/full}{full} \(\mathbf{Q}\). The fitting results for block-diagonal version is a bit better, because the underlying true \(\mathbf{Q}\) is diagonal. In the following part, I only show results from block-diagonal fitting for labeled data. For unlabeled data, i.e. clustering, all the results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
%
%\subsection{Simulation 1: Generate from LDS Directly}
%In this simulation, there are 3 clusters with 10 neurons in each cluster. The dimension of latents in each cluster is 2. In the linear dynamics, the bias term \(\mathbf{b}\) is zero. There's no within-population interaction but has some weak between-population interactions. In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal. The details of simulation can be found in the first section of the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{simulation 1}.
%
%\subsubsection{Labeled Data: No Clustering}
%The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{here}. 
%
%Here, I first compare 10k iterations for three methods (consider high dimension of \(\mathbf{x}_t\)): (1) both \(\mathbf{x}_t\) and loading (\(\mathbf{d}\) and \(\mathbf{C}\)) are updated by normal approximation. (2) update \(\mathbf{x}_t\) by normal approximation, but update loading by MH. (3) update both \(\mathbf{x}_t\) and loading by MH.
%
%The fitting results are average of iteration 5k to 10k.
%
%\textbf{Update \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by normal approximation}
%
%The trace plots are in Figure \ref{double N: trace}
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleN.png}}%
%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleN.png}}%
%	}
%	\caption{both Normal Approx., trace plots}
%	\label{double N: trace}
%\end{figure}
%
%Well, the chain hasn't achieve the stationary distribution. So there's no need to see further results. But to show the improvement and for comparison, I still show the fitting of overall firing rate, latents and dynamics in Figure \ref{double N: fit}.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleN.png}}%
%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleN.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleN.png}}%
%	}
%	\caption{both Normal Approx., fit}
%	\label{double N: fit}
%\end{figure}
%
%\textbf{Update \(\mathbf{x}_t\) by normal approximation and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
%
%By using the normal approximated posterior variance as the proposal variance, the acceptance rates are around 0.45 (Since there are 30 neurons, there are 30 acceptance rates).
%The trace plots are in Figure \ref{dc MH: trace}
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_dcMH.png}}%
%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_dcMH.png}}%
%	}
%	\caption{latent Normal Approx. + loading MH, trace plots}
%	\label{dc MH: trace}
%\end{figure}
%
%Looks much better, although still not perfect yet. The fitting of overall firing rate, latents and dynamics are in Figure \ref{dc MH: fit}.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_dcMH.png}}%
%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_dcMH.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_dcMH.png}}%
%	}
%	\caption{latent Normal Approx. + loading MH, fit}
%	\label{dc MH: fit}
%\end{figure}
%
%See, nearly perfect (especially) for latent.
%
%\textbf{Update both \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
%
%The acceptance rate (by using the optimal scalar) for latent is 0.6227, which is a bit high (but acceptable). Again, the acceptance rates for loading are around 0.45.
%
%The trace plots are in Figure \ref{double MH: trace}
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleMH.png}}%
%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleMH.png}}%
%	}
%	\caption{both MH, trace plots}
%	\label{double MH: trace}
%\end{figure}
%
%The trace for dynamics is terrible... Maybe this is caused by high dimension of latent (\(d=6000\) in this case).
%
%The fitting of overall firing rate, latents and dynamics are in Figure \ref{double MH: fit}.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleMH.png}}%
%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleMH.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleMH.png}}%
%	}
%	\caption{both MH, fit}
%	\label{double MH: fit}
%\end{figure}
%
%OK, the fitting sucks...
%
%One natural idea is to combine MH and normal approximation in latent: at first few steps, use normal approximation at conditional posterior mode to bring things into somewhere close to mode quickly. Then use MH to stabilize everything.
%
%In the following, I still sample loading by MH. But for latent, the first half iteration is sampled by normal approximation, while the remainder is sampled from MH.
%
%The red line marks when I switch the sampling method for \(\mathbf{x}_t\).
% 
%The trace plots are in Figure \ref{half MH: trace}
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_halfMH.png}}%
%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_halfMH.png}}%
%	}
%	\caption{half MH for latent, trace plots}
%	\label{half MH: trace}
%\end{figure}
%
%The fitting of overall firing rate, latents and dynamics are in Figure \ref{half MH: fit}.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_halfMH.png}}%
%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_halfMH.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_halfMH.png}}%
%	}
%	\caption{half MH for latent, fit}
%	\label{half MH: fit}
%\end{figure}
%
%Well, everything is stable now. But this implementation is too naive. Maybe use some more advanced adaptive sampling method?

%\begin{figure}[h!]
%	\centering
%	\includegraphics[width = .8\textwidth]{lds_blk_trace.png}
%	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
%	\label{lds norms}
%\end{figure}\\
%Some results are in Figure \ref{fig:LDS labeled}. These are means from iteration 1000 to 2000.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{lds_blk_FR.png}}%
%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{lds_blk_latent.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{lds_blk_A.png}}%
%	}
%	\caption{LDS sample with labels}
%	\label{fig:LDS labeled}
%\end{figure}
%
%Personally, I think the fitting is not bad and the mixing is fine (in previous version, we may need 10000 iterations to get a not bad result. But here 2000 iterations did even better than before).

%I also fit the model by assuming one cluster. Again, the fitting for overall mean firing rate is perfect. This is why it's important to make the loading, \(\mathbf{d}\) and \(\mathbf{C}\), also be cluster-dependent (loading within clusters are correlated).
%
%If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.
%
%
%\subsubsection{Unlabeled Data: Clustering}
%To give the full path of clustering, I show results in GIFs. All results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}. The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.
%
%Basically, in my current implementation, algorithms are good to merge clusters. However, generating new clusters is very hard... In other words, the newly generated cluster will usually not be sampled.
%
%\textbf{This is a big problem}. This makes DPMM loses its power. \textbf{Fix that later}.
%
%\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
%In this simulation, there are again 3 clusters with 2 latents and 10 neurons in each. Now, the latents are generated directly without specifying the underlying linear dynamics of latents. However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal. The details of simulation can be found in the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/unspecifiedA_sample_blkDiag.m}{code}.
%
%\subsubsection{Labeled Data: No Clustering}
%As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster. The code can be found \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/unspecifiedA_sample_blkDiag.m}{here}.
%
%
%\textbf{I have deleted the old results to avoid confusion. I will post the new results here later.}

%This time I ran 1000 iterations. The trace of the samples is shown in Figure \ref{noA norms}.
%\begin{figure}[h!]
%	\centering
%	\includegraphics[width = .8\textwidth]{noA_blk_trace.png}
%	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
%	\label{noA norms}
%\end{figure}\\
%
%Some results are in Figure \ref{fig:noA labeled}. These are means from iteration 500 to 1000.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{noA_blk_FR.png}}%
%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{noA_blk_latent.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{noA_blk_A.png}}%
%	}
%	\caption{LDS sample with labels}
%	\label{fig:noA labeled}
%\end{figure}
%
%This time the mixing of loading (\(\mathbf{d}\) and \(\mathbf{C}\)) is not fine. That reminds me that \textbf{I may need to make loading within cluster related by MV-(Poisson)-GLM directly.}

%\subsubsection{Unlabeled Data: Clustering}
%\textbf{The clustering results are old. Update when the MFM is implemented}
%
%Again, all results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}. The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.
%
%Besides fitting 3 clusters with 10 neurons each, I further fit a larger scale example (50 neurons for each cluster).

@@

\subsection{Constraints}
An affine transformation contains translation and linear transformation.

\subsubsection{Translation}
There are two strategies: (1) just ignore \(\mathbf{d}\) and let \(\mathbf{Cx}_{t}\)
absorb all the intercepts; (2) center \(\mathbf{x}_t\) such that mean for each
row of \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0. However, because of the block-diagonal constraints on $\mathbf{M}$, the strategy (1) may be problematic. Moreover, when $\mathbf{d}$ is non-zero, the \(\mathbf{x}_t\) will looks ugly. Another benefit for (2) is that we can use $\mathbf{d}$ to help clustering. However, the constraint on translation cannot resolve problems from linear transformation, since any linear transformation will preserve the zero row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\).

\subsubsection{Linear Transformation}
To mimic PCA, one natural way is to make the loading orthogonal. Again, I'm considering two strategies: (1) make
\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\); (2) make \( (\mathbf{C}'^{(1)},\cdots, \mathbf{C}'^{(J)})
\begin{pmatrix}
	\mathbf{C}^{(1)} \\
	\vdots \\
	\mathbf{C}^{(J)} \\
\end{pmatrix} = \mathbf{I}_{p}\). Personally, I prefer the first one, for both computational and clustering reasons. Computationally, if I do things cluster-wise, I can easily use parallel when dimension becomes large.

\textbf{However, this is not enough!} Consider \(\mathbf{x}_{t}^{*} = \mathbf{Ux}_{t}\), to
preserve the orthogonality and block diagonal on \(\mathbf{C}\), \(\mathbf{U}\) must be
block diagonal and each diagonal block \(\mathbf{U}^{(j)}\) is an orthogonal matrix (need to be the same for constraint (2)). More explicitly,

@@

\textbf{The above argument reminds me that I may need one more tiny constraint to solve the sign flipping issue. Do it later.}

In summary, we need add three constraints: (1) translation of \(\mathbf{x}_{t}\),
(2) orthogonality of loading and (3) diagonality of linear dynamics. In the following implementation, I tried 3 sets of constraints:

@@

Although, constraint (1) (diagonal $\mathbf{Q}$) is valid in theory, it doesn't work well in practice. Since elements $\mathbf{Q}$ is usually very small (~$10^{-3} ot 10^{-5}$).
Therefore, I will discard this constraint later.

\subsection{Constraint Implementation}

\subsubsection{Translation}
The constraint on translation is easy, just row-center the sample of \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) in each iteration. We can get the transformed full conditional distribution after some algebra. Let \(\mathbf{X} = \left( \mathbf{x}_{1},\ldots,\mathbf{x}_{T} \right) = \begin{pmatrix}
	\mathbf{X}_{(1)}^{'} \\
	\vdots \\
	\mathbf{X}_{(Jp)}^{'} \\
\end{pmatrix}\), where \(\mathbf{x}_{(k)} \in \mathbb{R}^{T}\) is the k-th row vector.
Then the k-th row mean is
\({\overline{\mathbf{x}}}_{(k)} = \frac{1}{T}\mathbf{x}_{(k)}^{'}\mathbf{1}_{T}\). The unconstraint full conditional distribution of \(vec(\mathbf{X})\) is derived in previous section. After
normalization in each iteration, we are actually sampling from
\(P(vec(\mathbf{X}) - \mathbf{1}_{T}\bigotimes\begin{pmatrix}
	{\overline{\mathbf{x}}}_{(1)} \\
	\vdots \\
	{\overline{\mathbf{x}}}_{(Jp)} \\
\end{pmatrix}|\mathbf{Y},\ldots)\). Notice that,
\[\mathbf{1}_{T}\bigotimes\begin{pmatrix}
	{\overline{\mathbf{x}}}_{(1)} \\
	\vdots \\
	{\overline{\mathbf{x}}}_{(Jp)} \\
\end{pmatrix} = \mathbf{1}_{T}\bigotimes\frac{1}{T}\mathbf{X1}_{T} = \frac{1}{T}vec\left( \mathbf{1X}_{T}\mathbf{1}_{T}^{'} \right) = \frac{\mathbf{J}_{T}\bigotimes \mathbf{I}_{Jp}}{T}vec(\mathbf{X})\]
, then we are just sampling from full conditional distribution of
\((\mathbf{I}_{JpT} - \frac{\mathbf{J}_{T}\bigotimes \mathbf{I}_{Jp}}{T})vec(\mathbf{X})\)

\subsubsection{Linear Transformation}
The constraint on orthogonality is kind of tricky. Here, I just transformed \(\mathbf{C}^{(j)}\) be
\(\mathbf{C}^{(j)} = \mathbf{K}^{(j)}\left( \mathbf{K}^{'(j)}\mathbf{K}^{(j)} \right)^{- 1/2}\). Therefore,
we can just sample \(\mathbf{K}^{(j)}\) without any constrained. Since
\(\mathbf{K}^{(j)}\) has the same structure as \(\mathbf{C}^{(j)}\), we can put some
hierarchical prior to model \(\mathbf{C}^{(j)}\). Specially, when
\(\mathbf{K}^{(j)} \sim MN_{n,p}(\mathbf{0},\bm{\Sigma}^{(j)},\mathbf{I})\) (\(MN_{n,p}\) means a matrix
normal distribution), \(\mathbf{C}^{(j)} \sim MACG(\bm{\Sigma}^{(j)})\).
\(MACG(\bm{\Sigma})\) means the matrix angular central Gaussian distribution,
with density
\(f_{\mathbf{X}}(\mathbf{X}) = |\bm{\Sigma}\left. \  \right|^{- p/2}{|\mathbf{X}'\Sigma^{- 1}\mathbf{X}|}^{- n/2}\).
When \(\bm{\Sigma}^{(j)} = \mathbf{I}_{n}\) or \(n = p\), \(\mathbf{C}^{(j)}\) is just
uniformly distributed on the Stiefel manifold
\(S(p,n) = \left\{ \mathbf{C} \in \mathbb{R}^{n \times p} \right|\mathbf{C}'\mathbf{C} = \mathbf{I}_{p}\}\).

\textbf{This reminds me to fix the problem when \(n < p\)}, which may happen when doing clustering. One easy solution is just use the constraint
\( (\mathbf{C}'^{(1)},\cdots, \mathbf{C}'^{(J)})
\begin{pmatrix}
	\mathbf{C}^{(1)} \\
	\vdots \\
	\mathbf{C}^{(J)} \\
\end{pmatrix} = \mathbf{I}_{p}\), or just merge clusters with \(n < p\) to other
clusters. Fix that.

@@

\subsection{Simulation}
The simulation example is as before, and all the code can be found in \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{this folder}.

>>>>

@@

@@

\subsection{Constraints}

%\subsubsection{constraint on $\bm{x}_t$ only}
\subsubsection{Constraint on latent vectors $\mathbf{x}_t$ only}
The easiest way to constraint is to normalize or standardize each row of \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\). Let \(\mathbf{X} = \left( \mathbf{x}_{1},\ldots,\mathbf{x}_{T} \right) = \begin{pmatrix}
	\mathbf{X}_{(1)}^{'} \\
	\vdots \\
	\mathbf{X}_{(Jp)}^{'} \\
\end{pmatrix}\), where \(\mathbf{x}_{(k)} \in \mathbb{R}^{T}\) is the k-th row vector. Then,
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Normalization: \(\mathbf{x}_{(k)}^* = \frac{\mathbf{x}_{(k)}-\min \mathbf{x}_{(k)}}{\max \mathbf{x}_{(k)} - \min \mathbf{x}_{(k)}}\)
	\item
	Standardization: \(\mathbf{x}_{(k)}^* = \frac{\mathbf{x}_{(k)} - \bar{\mathbf{x}}_{(k)}}{S(\mathbf{x}_{(k)})}\)
\end{enumerate}
The purpose of this constraint is to let $\bm{d}$ capture the translation and $\bm{C}$ capture the scaling.

After normalization/ standardization, the solution should be unique.

\subsubsection{Joint constraint on $\mathbf{x}_t$, $\mathbf{C}$ and linear dynamics}

\textbf{This is perfect when cluster is known} (more robust than constraints on $\bm{x}_t$ only), but when doing clusters, this is not very useful... Since the constraint depends on the number of elements in the cluster, but the number of elements will change when doing clustering.

See details in the appendix.

\subsection{Constraint Implementation}
The constraint for normalization and standardization are fairly easy: just do transformation of the MCMC sample in each iteration.

Normalization and standardization are both linear transformation, so after some algebra, we can even write the full conditional distribution explicitly.

For example, let's write out the linear transformation for centering. Denote the k-th row mean as
\({\overline{\mathbf{x}}}_{(k)} = \frac{1}{T}\mathbf{x}_{(k)}^{'}\mathbf{1}_{T}\). The unconstraint full conditional distribution of \(vec(\mathbf{X})\) is derived in previous section. After
normalization in each iteration, we are actually sampling from
\(P(vec(\mathbf{X}) - \mathbf{1}_{T}\bigotimes\begin{pmatrix}
	{\overline{\mathbf{x}}}_{(1)} \\
	\vdots \\
	{\overline{\mathbf{x}}}_{(Jp)} \\
\end{pmatrix}|\mathbf{Y},\ldots)\). Notice that,
\[\mathbf{1}_{T}\bigotimes\begin{pmatrix}
	{\overline{\mathbf{x}}}_{(1)} \\
	\vdots \\
	{\overline{\mathbf{x}}}_{(Jp)} \\
\end{pmatrix} = \mathbf{1}_{T}\bigotimes\frac{1}{T}\mathbf{X1}_{T} = \frac{1}{T}vec\left( \mathbf{1X}_{T}\mathbf{1}_{T}^{'} \right) = \frac{\mathbf{J}_{T}\bigotimes \mathbf{I}_{Jp}}{T}vec(\mathbf{X})\]
, then we are just sampling from full conditional distribution of
\((\mathbf{I}_{JpT} - \frac{\mathbf{J}_{T}\bigotimes \mathbf{I}_{Jp}}{T})vec(\mathbf{X})\)

To help the convergence, we may further add constraint on linear dynamics. Most of them can be fairly implemented, by diagonal $\mathbf{A}$ with full $\mathbf{Q}$ need some algebra. See details in the appendix.

\subsection{Simulation}
Here, I ran both normalization and standardization for 10,000 iterations. Although pure normalization or standardization should be enough to ensure convergence, further constraints in linear dynamics may help convergence.
Several constraints in linear dynamics (i.e. $\bm{A}$ and $\bm{Q}$) are also implemented:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	No constraints in $\bm{A}$ or $\bm{Q}$
	\item
	No constraints in $\bm{A}$, but $\bm{Q}$ is block-diagonal.
	\item
	No constraints in $\bm{A}$, but $\bm{Q}$ is diagonal.
	\item
	$\bm{A}$ is diagonal, but no constraints in $\bm{Q}$
	(See details in appendix).
	\item
	Both $\bm{A}$ and $\bm{Q}$ are diagonal.
\end{enumerate}

By limited experiments, it seems when $\bm{A}$ is diagonal, the convergence \& mixing is best. If we put no constraint on $\bm{A}$, then putting no constraint on $\bm{Q}$ is the best. Here I show the 2-norm/ Frobenius norm of \(\left\{ \mathbf{d},\mathbf{C},\mathbf{X},\mathbf{b},\mathbf{A},\mathbf{Q} \right\}\) for 4 constraints combos:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Normalization + both $\bm{A}$ and $\bm{Q}$ are diagonal.
	\item
	Normalization + $\bm{A}$ diagonal but $\bm{Q}$ full.
	\item
	Standardization + both $\bm{A}$ and $\bm{Q}$ are diagonal.
	\item
	Standardization + $\bm{A}$ diagonal but $\bm{Q}$ full.
\end{enumerate}

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[normalization + $\bm{A}$ and $\bm{Q}$ diagonal]{\label{fig:a}\includegraphics[width=0.75\textwidth]{norm_AQ_diag_dCX_trace_full.png}}%
		\subfigure[normalization + $\bm{A}$ diagonal and $\bm{Q}$ full]{\label{fig:b}\includegraphics[width=0.75\textwidth]{norm_A_diag_dCX_trace_full.png}}%
	}
	\caption{normalization, full trace of $\bm{d}$, $\bm{C}$ and $\bm{X}$}
	\label{full trace, normal, obs}
\end{figure}

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[standardization + $\bm{A}$ and $\bm{Q}$ diagonal]{\label{fig:a}\includegraphics[width=0.75\textwidth]{std_AQ_diag_dCX_trace_full.png}}%
		\subfigure[standardization + $\bm{A}$ diagonal and $\bm{Q}$ full]{\label{fig:b}\includegraphics[width=0.75\textwidth]{std_A_diag_dCX_trace_full.png}}%
	}
	\caption{standardization, full trace of $\bm{d}$, $\bm{C}$ and $\bm{X}$}
	\label{full trace, std, obs}
\end{figure}

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[normalization + $\bm{A}$ and $\bm{Q}$ diagonal]{\label{fig:a}\includegraphics[width=0.75\textwidth]{norm_AQ_diag_bAQ_trace_full.png}}%
		\subfigure[normalization + $\bm{A}$ and $\bm{Q}$ full]{\label{fig:b}\includegraphics[width=0.75\textwidth]{norm_A_diag_bAQ_trace_full.png}}%
	}
	\caption{normalization, full trace of $\bm{b}$, $\bm{A}$ diagonal and $\bm{Q}$}
	\label{full trace, normal, dynamics}
\end{figure}

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[standardization + $\bm{A}$ and $\bm{Q}$ diagonal]{\label{fig:a}\includegraphics[width=0.75\textwidth]{std_AQ_diag_bAQ_trace_full.png}}%
		\subfigure[standardization + $\bm{A}$ and $\bm{Q}$ full]{\label{fig:b}\includegraphics[width=0.75\textwidth]{std_A_diag_bAQ_trace_full.png}}%
	}
	\caption{standardization, full trace of $\bm{b}$, $\bm{A}$ diagonal and $\bm{Q}$}
	\label{full trace, std, dynamics}
\end{figure}

\clearpage
\section{Appendix}

\subsection{Joint constraint on $\mathbf{x}_t$, $\mathbf{C}$ and linear dynamics}
For $\mathbf{x}_t$, we can row center it as previous. But instead of doing scaling on $\bm{x}_t$, we can put the scaling to the loading $\mathbf{C}$.

To mimic PCA, one natural way is to make the loading orthogonal. Here, I simply make \(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\).

\textbf{However, this is not enough!} Consider \(\mathbf{x}_{t}^{*} = \mathbf{Ux}_{t}\), to
preserve the orthogonality and block diagonal on \(\mathbf{C}\), \(\mathbf{U}\) must be
block diagonal and each diagonal block \(\mathbf{U}^{(j)}\) is an orthogonal matrix. More explicitly,

@@

In summary, we need add three constraints if do things jointly: (1) translation of \(\mathbf{x}_{t}\),
(2) orthogonality of loading and (3) diagonality of linear dynamics. In the following implementation, I tried 3 sets of constraints:

@@

Although, constraint (1) (diagonal $\mathbf{Q}$) is valid in theory, it doesn't work well in practice. Since elements $\mathbf{Q}$ is usually very small (~$10^{-3}$ to $10^{-5}$).
Therefore, I will discard this constraint later.

\subsection{sample the orthogonal loading}
The constraint on orthogonality is kind of tricky. Here, I just transformed \(\mathbf{C}^{(j)}\) be
\(\mathbf{C}^{(j)} = \mathbf{K}^{(j)}\left( \mathbf{K}^{'(j)}\mathbf{K}^{(j)} \right)^{- 1/2}\). Therefore,
we can just sample \(\mathbf{K}^{(j)}\) without any constrained. Since
\(\mathbf{K}^{(j)}\) has the same structure as \(\mathbf{C}^{(j)}\), we can put some
hierarchical prior to model \(\mathbf{C}^{(j)}\). Specially, when
\(\mathbf{K}^{(j)} \sim MN_{n,p}(\mathbf{0},\bm{\Sigma}^{(j)},\mathbf{I})\) (\(MN_{n,p}\) means a matrix
normal distribution), \(\mathbf{C}^{(j)} \sim MACG(\bm{\Sigma}^{(j)})\).
\(MACG(\bm{\Sigma})\) means the matrix angular central Gaussian distribution,
with density
\(f_{\mathbf{X}}(\mathbf{X}) = |\bm{\Sigma}\left. \  \right|^{- p/2}{|\mathbf{X}'\Sigma^{- 1}\mathbf{X}|}^{- n/2}\).
When \(\bm{\Sigma}^{(j)} = \mathbf{I}_{n}\) or \(n = p\), \(\mathbf{C}^{(j)}\) is just
uniformly distributed on the Stiefel manifold
\(S(p,n) = \left\{ \mathbf{C} \in \mathbb{R}^{n \times p} \right|\mathbf{C}'\mathbf{C} = \mathbf{I}_{p}\}\).

@@

\section{Simulation on joint constraint}
The simulation example is as before, and all the code can be found in \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{this folder}.
